Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph

### 2015-03-10 curvature.tex
A sub-pass through tex. cc @cwhidden.

\newcommand{\RR}{\mathbb R}
\newcommand{\TT}{\mathcal T}

@@

\subsection{Computing curvature values}
We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
We then used \dots \todo{ILP, citation, github reference} to compute curvatures for given pairs of trees.
This latter step would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
We thus directly computed curvature values for one representative pair from each such equivalance class, or tanglegram. \todo{cite} \todo{indicate that a paper on tanglegrams is in preparation}

\subsection{Simulating random walks on the rSPR graph}
\label{sec:random_walks}
We next explored the correspondence between curvature and graph effects.
To do so, we first simulated Metropolis-Hastings random walks on the rSPR graphs with up to seven leaves.
The uniform random walk moves to a neighboring vertex in the graph uniformly at random.
However, this makes the walk more likely to sample higher degree vertices.
In contrast, the Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The MH random walk is guaranteed to sample each tree uniformly at random and is therefore representative of a phylogenetic MCMC program sampling trees under a uniform prior.

To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be precomputed.
Naively, this would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree and then pick one uniformly at random.
To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves such that an arbitrary SPR neighbor can be selected by its index in $\OhOf{n}$ time.
Song \todo{cite} developed a recursive formula for the degree of a tree and an explicit equation that works in terms of the distribution of heights of each node.
This formula does not group SPR moves that move a particularly subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.

Instead, we consider the distribution of SPR moves in terms of the number of nodes contained within a subtree.
Recall that a tree with $n$ leaves has $2n-1$ nodes.
Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
However, some of these moves will result in the same neighboring tree as other SPR moves.
In particular:
\begin{enumerate}
\item Moving $R$ to its sibling results in the same tree, not a neighboring tree,
\item Moving $R$ to its parent results in the same tree,
\item Moving $R$ to its grandparent is the same as moving its aunt to its sibling
\item Moving $R$ to its aunt is the same as moving its aunt to R.
\end{enumerate}
We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
In this manner, we achieve an alternative solution for computing the degree of a tree:
\todo{equation}
In particular, this formulation groups moves that move the same subtree.
We can thus apply the following algorithm to select a neighbor uniformly at random:

\begin{enumerate}[label={\arabic*}.]
\item	compute the degree
\item pick a random number $r$ from $1$ to the degree
\item uniquely order the tree by the smallest alphanumeric label of descendants
\item preorder the nodes of $R$
\item for each tree node $R$ in order and while $r > 0$
	\begin{enumerate}
		\item decrease $r$ by $2n-1-x-4$ or $2n-1-x-2$
		\item if $r < 0$, pick the $\abs{r}$ valid move of $R$ in order
	\end{enumerate}
\item apply the corresponding move and return the corresponding tree
\end{enumerate}

@@

We continue to explore the effect of biased random walks by considering the curvature of Metropolis Hastings random walks.
The Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
We now bound the difference between the curvature of uniform random walks $\curvature{S,T}$, and that of MH random walks, $\curvature[\mathrm{MH}]{S,T}$.

\begin{lemma}
	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
	\le \curvature[MH]{T,S}
	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }$$
	$$\curvature{T,S} - \frac{1}{6}
	\le \curvature[MH]{T,S}
	\le \curvature{T,S} + \frac{1}{6}.$$
\end{lemma}
\begin{proof}
	We first prove the lower bound.
	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
	Thus, the hastings ratio is always $\ge \frac{5}{6}$.
	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$.
	We construct an upper bound on $W_1'(T,S)$ for the MH random walk by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, $T$ to a neighbor of $S$, or from $T$'s neighbors to $S$.
	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
	Moreover, the affected mass must be moved through at most two additional trees.
	Then, $W_1' \le W_1 + \frac{2}{6}$.
	We now have:
	\begin{align*}
		\curvature[\mathrm{MH}]{T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}
	\end{align*}

@@

SAGE and GAP4 software

>>>>

\newcommand{\RR}{\mathbb R}
\newcommand{\TT}{\mathcal T}
\newcommand{\MH}{\operatorname{MH}}

@@

\subsection{Computing curvature values}
We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
We first used \todo{\dots} to compute the rSPR treespaces for four to seven leaves.
This required \dots time.
We then used linear programming to compute curvatures for given pairs of trees \cite{gricci}.
This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} (enumeration methods described in a manuscript in preparation).

\subsection{Simulating random walks on the rSPR graph}
\label{sec:random_walks}
We next explored the correspondence between curvature and graph effects.
To do so, we first simulated Metropolis-Hastings random walks on the rSPR graphs with up to seven leaves.
The uniform random walk moves to a neighboring vertex in the graph uniformly at random.
However, this makes the walk more likely to sample higher degree vertices.
In contrast, the Metropolis Hastings (MH) random walk with constant likelihood function proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The MH random walk is guaranteed to sample each tree uniformly at random and is therefore representative of a phylogenetic MCMC program sampling trees under a uniform prior.

To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built.
A na\"ive approach would require $\OhOf{n^3}$ time---$OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves such that an arbitrary SPR neighbor as described in the next paragraph.
Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
We note that the recursive formula of Song \cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.

Instead, we consider the distribution of SPR moves in terms of the number of nodes contained within a subtree.
Recall that a tree with $n$ leaves has $2n-1$ total nodes.
Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
However, some of these moves will result in the same neighboring tree as other SPR moves.
In particular, where ``sibling edge'' means the edge connecting the sibling subtree to the rest of the tree, etc.:
\begin{enumerate}
\item Moving $R$ to its sibling edge results in the same tree, not a neighboring tree,
\item Moving $R$ to its parent edge results in the same tree,
\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge
\item Moving $R$ to its aunt edge is the same as moving its aunt to $R$'s edge.
\end{enumerate}
\todo{How do we know that there aren't any other symmetries?}
We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
In this manner, we achieve an alternative solution for computing the degree of a tree in the SPR graph:
\todo{equation}
In particular, this formulation groups moves that move the same subtree.
We can thus apply the following algorithm to select a neighbor uniformly at random:

\begin{enumerate}[label={\arabic*}.]
\item	compute the degree
\item pick a random number $r$ from $1$ to the degree
\item uniquely order the tree by the smallest alphanumeric label of descendants
\item preorder the nodes of $R$ \todo{preorder how?}
\item for each tree node $R$ in order and while $r > 0$
	\begin{enumerate}
		\item decrease $r$ by $2n-1-x-4$ or $2n-1-x-2$
		\item if $r < 0$, pick the $\abs{r}$ valid move of $R$ in order
	\end{enumerate}
\item apply the corresponding move and return the corresponding tree
\end{enumerate}

@@

We continue to explore the effect of biased random walks by considering the curvature of Metropolis Hastings random walks.
The Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
We now bound the difference between the curvature of uniform random walks $\curvature{S,T}$, and that of MH random walks, $\curvature[\MH]{S,T}$.

\begin{lemma}
	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
	\le \curvature[\MH]{T,S}
	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }$$
	$$\curvature{T,S} - \frac{1}{6}
	\le \curvature[\MH]{T,S}
	\le \curvature{T,S} + \frac{1}{6}.$$
\end{lemma}
\begin{proof}
	We first prove the lower bound.
	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
	Thus, the hastings ratio is always $\ge \frac{5}{6}$.
	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$.
	We construct an upper bound on $W_1'(T,S)$ for the MH random walk by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, $T$ to a neighbor of $S$, or from $T$'s neighbors to $S$.
	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
	Moreover, the affected mass must be moved through at most two additional trees.
	Then, $W_1' \le W_1 + \frac{2}{6}$.
	We now have:
	\begin{align*}
		\curvature[\MH]{T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}
	\end{align*}

@@

SAGE and GAP4 software
Alexander Hulpke
